Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-03-25 paper.qmd
image + caption updates

Uncertainty visualisation is a somewhat new subfield of visualisation, with early papers that specifically reference an "uncertainty visualisation" popping up in the late 90s and early 2000s (*Cite Gap 1*). Despite the youth of the field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves (*Cite Gap 2*). Early mentions of "uncertainty visualisation" appear in computer science papers discussing uncertainty visualisation as a subfield of geospatial information visualisation, specifically for the many uncertainties that come with satellite images (*Cite Gap 1*). In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts and is no longer bound by geospatial information (*Cite Gap 1*).

@@

Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision. Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3*). This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3.5*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

@@

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects. This is obvious in practice, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a]. That being said, these taxonomies can hint towards what is important when we think about uncertainty. The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021]. The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision". It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*). It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.

@@

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty.](sketchiness.png){#fig-sketchy}

@@

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

@@

### Definitions of "uncertainty visualisation"
A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 10*). @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.

@@

Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution"). This conceptualisation could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous. However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty. Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important. (*Cite Gap 10*)

@@

![This visualisation is here as a placeholder. I am going to replace it with an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014 ](paperscreenshots/collage2.jpg){#fig-location}

@@

With this in mind, let us turn to the concept of "contextual" information. For the purposes of this review, we will consider contextual information to be anything that makes it easy for the audience to connect the data from a probabilistic kind to its real world parallel, but is largely irrelevant in the abstract probability space. A key example of this is spatial information, such as borders, road locations, terrains, etc. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model. @fig-spatial shows the process of boiling information down into data, visualising it, and including contextual information to improve our scientific reasoning. The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:

@@

![Data pipeline with visualisations to show why we are ignoring contextual information for the time being. The visualisations that depict the data are similar to typical statistical graphics. The purpose of the spatial information is to provide context to the information we extract from the statistical aspect of the graphic.](spatialinfo.jpeg){#fig-spatial}

@@

![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction.](bustime.png){#fig-bus}

@@

In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain. This is not only the case for uncertainty data. @Slingsby2023 discussed using a gridded glyph map for spatial modelling, however the statistical elements are identical to a fluctuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011. The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background. @Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework. This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endeavour, but these plots seem to be have been developed unaware of the existing statistical framework. When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-discipline.

```{r}
#| output: asis
#| echo: false
img_files <- fs::dir_ls("prodplots", glob="*.png")
cat("::: {layout-ncol=2}\n",
    glue::glue("![]({img_files})\n\n\n"),
    ":::",
    sep = ""
)
```

The focus on contextual information has also created a rift in exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.  
Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 11*). These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.

@@

### Information asymmetry in evaluated plots
Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. *Other examples* (*Cite Gap 13*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminiscent of our problems with the definition of "uncertainty".

@@

![The grammar of graphics data analysis pipeline from @Leland2005. the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}

@@

Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation. @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations". They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot). @Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean. They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution. @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations". They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately. *Other examples* (*Cite Gap 14*).

@@

It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research. Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost. To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half. If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale. @Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987]. Similar results to this occur over and over again in the uncertainty visualisation literature. @Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual. @Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. *Other examples* (*Cite Gap 15*).

@@

@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location. Their payment scheme, which paid out $0.10 for a correct choice (i.e. staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%. If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*. The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another. Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation. (*Cite Gap 16*)

@@

![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation. ](furthestright.png){#fig-right}

@@

::: {#fig-maps}

![](mappals/img1.png)

>>>>

Uncertainty visualisation is a somewhat new subfield of visualisation. Early papers that specifically reference "uncertainty visualisation" start popping up in the late 80s [@Ibrekk1987], with papers in the early 90s (specifically in the area of geospatial information visualisation) recgonising it as a necessary subfield of visualisation [@MacEachren1992; Carr1992]. Despite the youth of the specific sub-field, commonly used uncertainty visualisations such as a box plot or histogram have been around for almost as long as statistical graphics themselves. In the modern day, uncertainty visualisation has exploded into its own field that applies to a wide range of contexts.

@@

Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision. Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3: Organise paper definitions*). This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3: Organise paper definitions*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4: Organise uncertainty quantification*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods don't even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 4: Organise uncertainty quantification*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

@@

## Taxonomy definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7: Uncertainty Taxonomies*). In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertainty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects. This is obvious in practice, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a]. That being said, these taxonomies can hint towards what is important when we think about uncertainty. The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021]. The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision". It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]. It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.

@@

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty. These graphic were made without concern as to what (i.e. which statistic) in the image is supposed to be uncertain. This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent. For example, it is unclear how you can be uncertain about a train line (S5).](sketchiness.png){#fig-sketchy width=50%}

@@

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisation, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 4: Organise uncertainty quantification*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

@@

### Definitions of "uncertainty visualisation"
A large drive of the field of uncertainty visualisation, has been the idea that "uncertainty visualisation" (defined as one word) is somehow different to "uncertainty" "visualisation" (separately defined as two words). When "uncertainty visualisation" is defined as a single word we lose some nuance in our understanding of the driving force behind some issues, and problems with visualisation and problems with uncertainty become conflated. This appears in the literature as a mountain of seemingly conflicting statements about what is, or is not, an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap 8: Bar/pie uncertainty graphics*). @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. These conflicting ideas and lacking definition of what is means for a visualisation to be an "uncertainty visualisation" is a cornerstone of some clearly questionable results within this field.

@@

Some taxonomies of uncertainty highlight how confusing most authors find the concept, and how difficult it is to articulate the rules of an uncertainty visualisation when uncertainty itself if not defined. @Potter2012 organised several existing uncertainty visualisations into groups based on the dimensionality of the data (1D, 2D, 3D, and No Dimension) and the dimensionality of the uncertainty (Scalar, Vector, Tensor).  However, because the term "PDF", a statistical object that describes a random variable that is typically a one dimensional function, is used to describe both the data and the uncertainty for all dimensions, it is hard to understand how this should work. @Grewal2021 created a taxonomy that mapped uncertainty visualisations to some point in a 2D space defined by the "domain expertise" and "continuum of discreteness" (that scaled from "point estimate" to "continuous distribution"). This conceptualisation could be considered a visual extension of the law of large numbers, because as a point estimate is a single sample, as the sample size gets larger the distribution will appear to be a continuous. However, assumes every point estimate is a sample from the same distribution (a mean is not technically a sample from a prediction interval, but it is a sample from the sampling distribution) and it conflates discreteness from a too small a number of observations, discreteness from a precision problem, discreteness because that is the true nature of the variable, and discreteness as a visualisation choice. @Griethe2006 organised uncertainty visualisations into two cases (1) a hypothesis test was preformed to confirm the validity of the visualisation and (2) the visualisation has uncertainty depicted. This conceptualization conflates hypothesis testing and confidence intervals, and while they are related concepts, a hypothesis test is not necessarily uncertainty. Of course the authors of these papers did not intend for the taxonomy to be a complete description of uncertainty visualisations, however these distinctions that are often ignored are subtly important. (*Cite Gap 9: Other uncertainty vis taxonomies*)

@@

![Planned visualisation: Add an R coded visualisation that shows the 5 considerations for showing uncertainty in a plot according to @Kinkeldey2014. One that has the consideration, and one that doesn't (try and keep as much the same between visuals as possible)](notice.jpeg){#fig-location width=50%}

@@

With this in mind, let us turn to the concept of "contextual" information. For the purposes of this review, we will consider contextual information to be anything that makes it easy for the audience to connect the data from a probabilistic kind to its real world parallel, but is largely irrelevant in the abstract probability space. A key example of this is spatial information, such as borders, road locations, terrains, etc. Spatial information is *highly relevant* when we translate from the abstract domain back into real world for communication. It can help provide reasons for *why* there may be specific characteristics in our data, however it is not required information for the model. @fig-datapipe shows the process of boiling information down into data, visualising it, and choosing whether or not to including contextual information to improve our scientific reasoning. The concept that contextual information is not the same as statistical information is also highlighted in *The Grammar of Graphics* which states:

@@

::: {#fig-datapipe layout-ncol=2}

![Australia example](spatialinfo.jpeg)

![USA example](spatialinfo2.jpeg)

Data pipeline with visualisations to show why we are ignoring contextual information in the case of uncertainty visualisation. The visualisations that depict the data are similar to typical statistical graphics. The Australia example depicts a heat map for temperature in Australia. The statistical level of this graphic is no different than a typical density plot that could be made with  exist with any data. The geographic information is merely used to provide real world context to the statistical message we extract from graphic, however that addition is of great importance. The USA example depicts a map showing the voting proportions in a US election. This map illustrates another example of a spatial graphic, however the location of the data is only loosely related to the real world geographic location. The graphic maintains readability because the proportion of votes matters as the US state level, and the number of people in a state and its proportion of votes are far more important for drawing inference on who won an election than specific spatial information. Notice: Both Graphic needs to be fixed up.
:::

@@

![The uncertainty visualisation used by @Kay2016 to show the uncertainty around a bus arrival time prediction. The data being a "time" estimate does not prohibit the authors from using a density plot, a visualisation method that is typically used for attributes. The time element is irrelevant because the viewer is concerned about their estimated wait time NOW, not how this estimate will change over time. The predictions relationship to any predictors is assumed to be irrelevant in this visualisation.](bustime.png){#fig-bus width=50%}

@@

In this sense we are not proposing that the current conceptualisation of data according to its physical properties is necessarily wrong, or that it leads to conclusions that are not necessarily specific to spatial or temporal data, and it may lead us to miss relevant information that is outside our domain. This is not only the case for uncertainty data. @Slingsby2023 discussed using a gridded glyph map for spatial modelling, however the statistical elements are identical to a fluctuation diagram depicting $f(infection|time, age, longitude, latitude)$ presented in @Wickham2011. A side by side depiction of the two graphics is displayed in @fig-prodplots. The plot by @Slingsby2023 is simply a special case of the graphic by @Wickham2011, where a separate visualisation is made for each time point, the continuous function $infection | age$ has replaced the categorical variable $happy$, a fade is added for the population size, and a map is shown in the background. @Slingsby2023 presents other maps in the paper, however they are also special cases of the product plots framework. This is not to say this extension is uninteresting, on the contrary we believe identifying cases where particular graphics are useful is a worthwhile endeavour, but these plots seem to be have been developed unaware of the existing statistical framework. When we separate the field of visualisation along lines that may not be of particular relevance, we increase the likelihood of "reinventing the wheel" in each sub-discipline.

::: {#fig-prodplots}

![Gridded glyphmap](prodplots/img1.png)
![Fluctuation diagram](prodplots/img2.png)

A side by side comparison of the gridded glyphmap and the fluctuation diagram. One is considered a spatial visualisation technique and the other is considered a normal statistical diagram. The product plot and the gridded glyph map are visualisations depicting the relationship $f(w|x,y,z)$ and the similarity in the two plots shows that this three (or 4) variable relationship is the key feature of statistical importance, not the spatial nature of the data. The limitations faced by spatial visualisation are often identical to the limitations faces by information visualisation and creating unnecessary divides within the field makes it more difficult to navigate and find relevant statistical visualisations. 
:::

The focus on contextual information has also created a rift in exactly *why* uncertainty visualisation is a field of interest. The two most common justifications are:
1) Uncertainty is fundamentally differently to other variables due to psychological heuristics involved in the interpretation of uncertainty (e.g. risk aversion).
2) Uncertainty is not of interest in of itself and is typically layered on at the end of a the visualisation pipeline, so uncertainty visualisation is the field of adding error information into already established graphics and *thus* uncertainty information is a high dimensional visualisation problem.  
(*Cite Gap 10: Reason uncertainty visualisation is different*)
Papers will often discuss uncertainty in relation to one of these motivating reasons, the evaluation experiments motivated by (1) often perform different visualisations of PDFs, while the papers motivated by (2) will focus on trying to impute uncertainty as error within one of the existing channels, such as colour (*Cite Gap 10*). These dual motivations that co-exist with uncertainty visualisation authors rarely mentioning the others, create confusion in the field as to its purpose. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata.

@@

### Information asymmetry in evaluated plots
Visualisation authors are almost unanimous in commenting that the "information" in two plots must be the same in order for the visual techniques to be compared [@Cleveland1984; @Kinkeldey2014] (*Cite gap 12: Equal information*). However, what makes two visualisations equal in "information" is not consistent and is regularly mentioned without it being clear exactly what this information is. This is easily illustrated by the wide array of comments made by visualisation authors when they explain why two representations are being compared in their evaluation study. @Ibrekk1987 displayed 6 PDFs in their experiment because "formally equivalent representations are often not psychologically equivalent", but do not explain why what makes a representation "formally equivalent" or why only one representation was presented for the 95% CI, box plot, and CDF. @Hofman2020 comments that "theoretically" the sampling distribution of the mean and the prediction interval of a new observation are equal "so long as one knows the sample size", but does not recognise the several assumptions and statistical knowledge that is required to compare the two. (*Cite Gap 13: examples of what is defined as "information"*). @Kinkeldey2014 adopts an existing definition also that suggests two graphics are informationally equivalent if all the information in one plot is inferable from the other and vice-versa, but adds that two plots are computationally equivalent if that information can be extracted from both plots with similar easy and speed. It is clear this lacking definition of "information" in a visual is reminiscent of our problems with the definition of "uncertainty".

@@

![(Notice: Need to replace with a more relevant data pipeline that includes all the uncertainty stuff and how uncertainty arises at each stage.) The grammar of graphics data analysis pipeline from @Leland2005. the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}

@@

Ultimately, this all points to confusion as to when two visualisation contain information asymmetry, a problem that becomes more pronounced in uncertainty visualisation. @Ibrekk1987 compared a 6 visualisations of a PDF,  a box plot, a CDF and a mean with a confidence interval on the basis that all of them were "uncertainty visualisations". They found that people are better at extracting the mean from a plot when they are shown a plot that contains a mean with a confidence interval than when they are shown a PDF, a box plot, or a CDF (where the mean could not be read off the plot). @Hofman2020 and @Zhang2022 compared prediction and sampling distributions because they are both "uncertainty" that is typically depicted around the mean. They found that people are better at answering questions about a prediction interval when shown a prediction interval instead of a sampling distribution. @Hullman2015 compared the static error bars and violin plots of the marginal distributions of two variables ($A$ and $B$) to an animated plot that depicted a single outcome of the joint distribution of $A$ and $B$ in each frame of an animation, a comparison that was justified because all plots were "uncertainty visualisations". They found that the plot that depicted outcomes from $P(A, B)$ where the distance between the two outcomes was equivalent to $a-b$ was better at answering the question "What is $P(B<A)$" than violin or error bar plots depicting $P(A)$ and $P(B)$ separately. (*Cite Gap 14: Examples of mathematically obvious results*).

@@

It is difficult to find examples of uncertainty visualisation experiments where the plots do contain the same information, however when they do, the results align with existing information visualisation research. Technically, a PDF and a mean with confidence intervals both have enough information to extract the mean of the distribution, however they both have a very different computational cost. To extract the mean using a PDF, a participant would need to identify the point along the x axis that splits the area under the curve in half. If a participant is provided with a mean with a confidence interval, extracting the average is a simple task of reading the position on an aligned scale. @Ibrekk1987 found that when asking participants for the "best estimate" (which they thought should be interpreted as the mean of the distribution) of a skewed distribution, participants provided the mean when given a mean with confidence intervals and the mode when given a PDF [@Ibrekk1987]. Similar results to this occur over and over again in the uncertainty visualisation literature. @Gschwandtnei2016 found that visualisations where the start time of an interval could literally be read off the plot (error bars, centred error bars, and ambiguation) performed better than the plots (accumulated probability, gradient, and violin) where the start time involved some guesswork because the drop off was gradual. @Cheong2016 found that participants were better at answering questions when they were explicitly given the relevant probability in text rather than when they needed to read it off a map. (*Cite Gap 15: Examples of replicated perceptual task experiments*).

@@

@Cheong2016 tested multiple different visual representations of uncertainty for representing the likelihood of a house being burned down based on its location. Their payment scheme, which paid out $0.10 for a correct choice (i.e. staying when the house was not burned down or leaving when the house was burned down) and 0 for an incorrect choice (i.e.leaving when the house didn't burn down or staying when the house burned down), meant participants were incentivised to base their entire leave/stay decision on whether or not the likelihood of a fire at their house is above or below 50%. If the participants *did* correctly identify the optimum strategy and answer accordingly, the "decision making experiment" was actually *just a value extraction experiment*. The decision making aspect ceased to matter, all that mattered was identifying if the probability of a fire was greater than 50%. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. This means it is likely that introducing a payment scheme or a utility function creates too much mental labour for the participants to answer "correctly", and they would rather stick with their own risk utility function that feels *natural* than go through the mental labour of adopting another. Setting up an experiment that requires participants to do laborious calculations or strenuous mental effort for a small marginal benefit is unlikely to result in successful participant responses because it is not in the "spirit" of visualisation. (*Cite Gap 16: Examples of value extraction decision making tasks*)

@@

![This shows the user interface for the experiment performed by @Hofmann2012. The question of "furthest to the right" is open to interpretation.  ](furthestright.png){#fig-right width=50%}

@@

::: {#fig-maps layout-ncol=2}
